---
title: "ST117 Individual DRAFT Written Report - Phase 2 Task 1 (Woodland Vegetation – Tree Measurements)"
author: "Marcus Cheng 5683200 (Report Pod MS-05)"
date: "2026-04-25" 
output: pdf_document
geometry: top=2cm, bottom=2cm
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)

wl <- read.csv("ECN_VW1.csv")

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# a)
*include an output of three rows*

```{r a}
wl_wide <- wl %>%
  filter(FIELDNAME %in% c("DIAMETER", "HEIGHT", "DISTANCE", "NUM_STEMS")) %>%
  pivot_wider(
    id_cols = c(SITECODE, SYEAR, TREEID, STEMID, TREE_SPEC),
    names_from = FIELDNAME,
    values_from = VALUE,
    values_fn = mean
  ) %>%
  select(SITECODE, SYEAR, TREE_SPEC, DIAMETER, HEIGHT, DISTANCE, NUM_STEMS)

head(wl_wide, 3)
```

# b)
*Limit: 50 words + 1 table.*

```{r b}
model1 <- lm(HEIGHT ~ DIAMETER, data = wl_wide)

kable(summary(model1)$coefficients, digits = 3)
```
Diameter is chosen as the predictor and height as the response, since taller trees generally have wider trunks. The model is statistically significant with p < 2.2e-16 and a slope of 0.109. However, R-squared is low (0.15), showing that diameter does not explain much of the variation in height.

\newpage

# c)
*Limit: 400 words + 10 figures + 5 tables*

```{r c}
T05 <- wl_wide %>% filter(SITECODE == "T05")
T06 <- wl_wide %>% filter(SITECODE == "T06")
T08 <- wl_wide %>% filter(SITECODE == "T08")
T09 <- wl_wide %>% filter(SITECODE == "T09")

model_T05 <- lm(HEIGHT ~ DIAMETER, data = T05)
model_T06 <- lm(HEIGHT ~ DIAMETER, data = T06)
model_T08 <- lm(HEIGHT ~ DIAMETER, data = T08)
model_T09 <- lm(HEIGHT ~ DIAMETER, data = T09)

ggplot(wl_wide, aes(x = DIAMETER, y = HEIGHT)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Overall: Height vs Diameter", x = "Diameter", y = "Height")

ggplot(T05, aes(x = DIAMETER, y = HEIGHT)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  labs(title = "T05: Height vs Diameter", x = "Diameter", y = "Height")

ggplot(T06, aes(x = DIAMETER, y = HEIGHT)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "T06: Height vs Diameter", x = "Diameter", y = "Height")

ggplot(T08, aes(x = DIAMETER, y = HEIGHT)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "T08: Height vs Diameter", x = "Diameter", y = "Height")

ggplot(T09, aes(x = DIAMETER, y = HEIGHT)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "T09: Height vs Diameter", x = "Diameter", y = "Height")

plot(model1, which = 1, main = "Overall: Residuals vs Fitted")
plot(model_T05, which = 1, main = "T05: Residuals vs Fitted")
plot(model_T06, which = 1, main = "T06: Residuals vs Fitted")
plot(model_T08, which = 1, main = "T08: Residuals vs Fitted")
plot(model_T09, which = 1, main = "T09: Residuals vs Fitted")

kable(summary(model_T05)$coefficients, digits = 3)
kable(summary(model_T06)$coefficients, digits = 3)
kable(summary(model_T08)$coefficients, digits = 3)
kable(summary(model_T09)$coefficients, digits = 3)

data.frame(
  Model = c("Overall", "T05", "T06", "T08", "T09"),
  R2 = c(summary(model1)$r.squared, summary(model_T05)$r.squared,
         summary(model_T06)$r.squared, summary(model_T08)$r.squared,
         summary(model_T09)$r.squared),
  Slope = c(coef(model1)[2], coef(model_T05)[2], coef(model_T06)[2],
            coef(model_T08)[2], coef(model_T09)[2])
) %>% kable(digits = 3)
```
There is a positive correlation between diameter and height in all models, since large diameter trees usually would be taller. But there seems to be some variance among all models in how well the relationship between diameter and height holds in different monitored sites.

T05 has the best fit, with a slope of 1.108, showing that diameter has significant effects on height. On the other hand, T06 has a moderate fit, while T08 has a low fit. Other factors may affect the relationship between diameter and height in these two areas.

In T09, there is almost no relationship between diameter and height. This may be because the area is more diverse with different species growing differently and have varied environmental factors.

When considering the overall model, the model does not fit well relative to individual models. By pooling data, site-level variance is averaged out. The residual plots of T08 and T09 are similar to the overall model, while those of T05 and T06 are not as similar.

# d)
*Limit: 300 words + 4 figures + 4 tables*

```{r d}
ggplot(T05, aes(x = DIAMETER, y = HEIGHT, colour = factor(TREE_SPEC))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "T05: Height vs Diameter by species", x = "Diameter", y = "Height", colour = "Species")

ggplot(T06, aes(x = DIAMETER, y = HEIGHT, colour = factor(TREE_SPEC))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "T06: Height vs Diameter by species", x = "Diameter", y = "Height", colour = "Species")

ggplot(T08, aes(x = DIAMETER, y = HEIGHT, colour = factor(TREE_SPEC))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "T08: Height vs Diameter by species", x = "Diameter", y = "Height", colour = "Species")

ggplot(T09, aes(x = DIAMETER, y = HEIGHT, colour = factor(TREE_SPEC))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "T09: Height vs Diameter by species", x = "Diameter", y = "Height", colour = "Species")

model_T05_sp <- lm(HEIGHT ~ DIAMETER + factor(TREE_SPEC), data = T05)
model_T06_sp <- lm(HEIGHT ~ DIAMETER + factor(TREE_SPEC), data = T06)
model_T08_sp <- lm(HEIGHT ~ DIAMETER + factor(TREE_SPEC), data = T08)
model_T09_sp <- lm(HEIGHT ~ DIAMETER + factor(TREE_SPEC), data = T09)

data.frame(
  Site = c("T05", "T06", "T08", "T09"),
  R2 = c(summary(model_T05)$r.squared, summary(model_T06)$r.squared,
         summary(model_T08)$r.squared, summary(model_T09)$r.squared),
  R2_species = c(summary(model_T05_sp)$r.squared, summary(model_T06_sp)$r.squared,
                 summary(model_T08_sp)$r.squared, summary(model_T09_sp)$r.squared)
) %>% kable(digits = 3)
```
One issue affecting all four locations is the fact that there are several types of trees with varying diameter-height slopes. A straight regression line that would be applicable for all species would not be ideal because different species have different diameters and heights.

For location T05, species 2600 shows higher growth than species 237 as indicated by a steeper slope while species 441 has only one value so it makes no sense to plot a regression line for it. Species 570 is the dominant species for T06 and determines the slope of the line. In T08, the number of species exceeds 30, thus, contributing to the high scatter of data around the line.

Incorporation of species into the model as a categorical variable will improve the fit of data on all locations. The values of R-squared increase from 0.84 to 0.94 in T05, 0.50 to 0.69 in T06, 0.38 to 0.77 in T08, and 0.025 to 0.13 in T09.
